---
title: "Arrays between R and Python"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Arrays between R and Python}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

zunpy reads and writes the files NumPy saves arrays in: `.npy`, one array,
and `.npz`, a ZIP archive of them. Python is not needed on either side of
R; the file is the interface.

```{r}
library(zunpy)
```

## Reading and writing

`npy_write()` and `npy_read()` take a path, a connection, or (for reading)
a URL. `npy_encode()` and `npy_decode()` do the same with raw vectors.

```{r}
f <- tempfile(fileext = ".npy")
x <- matrix(c(1.5, 2, 3, 4, 5, 6), 2)
npy_write(x, f)
identical(npy_read(f), x)
npy_header(f)
```

What zunpy writes is what `numpy.save()` writes for the same array, byte
for byte, header and padding included, so a file round-trips through
either side unchanged.

## Types

Each NumPy dtype reads as the R type that holds it exactly:

| NumPy | R |
| --- | --- |
| `b1` | logical |
| `i1`, `u1`, `i2`, `u2`, `i4` | integer |
| `u4`, `f2`, `f4`, `f8` | double |
| `i8`, `u8` | double, or `bit64::integer64` |
| `c8`, `c16` | complex |
| `S<n>`, `U<n>` | character (UTF-8) |
| `V<n>` | list of raw vectors |
| `M8[D]` | Date |
| `M8[s]` to `M8[ns]` | POSIXct, in UTC |
| `m8` | difftime |
| structured | data frame |

Writing goes the other way, from the R type: logical to `|b1`, integer to
`<i4`, double to `<f8`, complex to `<c16`, character to `<U<n>`, `Date` to
`<M8[D]`, `POSIXct` to `<M8[us]`. `dtype` chooses another for one call:

```{r}
npy_header(npy_encode(c(1, 2, 3), dtype = "<f4"))$descr
npy_header(npy_encode(c("a", "bb"), dtype = "|S"))$descr
```

### Exact or refused

A value is converted exactly or not at all. A 64-bit integer beyond 2^53
does not fit a double without rounding, so it is an error unless it is
read as `integer64`:

```{r, eval = requireNamespace("bit64", quietly = TRUE)}
big <- npy_encode(bit64::as.integer64("9007199254740993"))
try(npy_decode(big))
npy_decode(big, int64 = "integer64")
```

Narrowing on write refuses what does not fit, naming the element:

```{r}
try(npy_encode(c(1, 2, 300), dtype = "|u1"))
```

`NA` goes only where the type has a place for it. A double `NA` is a NaN
with R's own payload, so it survives a round trip; NumPy sees a NaN. An
integer `NA` is the bit pattern of `-2^31`, which is an ordinary value to
NumPy, so writing and reading it needs `na = "allow"`:

```{r}
try(npy_encode(c(1L, NA)))
npy_decode(npy_encode(c(1L, NA), na = "allow"), na = "allow")
```

## Memory order

NumPy stores arrays in C order (last index fastest) by default; R stores
them in Fortran order (first index fastest). zunpy reads either so that
`a[i, j]` in R is `a[i-1, j-1]` in Python, permuting a C-order file in one
pass. It writes R's own order, which costs nothing and which NumPy reads
as well; `order = "C"` writes C order for readers that need it.

```{r}
npy_header(npy_encode(x))$fortran_order
npy_header(npy_encode(x, order = "C"))$fortran_order
```

## Records and archives

A data frame is written as a record array, one field per column, and a
record array is read as a data frame:

```{r}
df <- data.frame(id = 1:3, score = c(0.5, NA, 2), name = c("a", "b", "c"))
npy_header(npy_encode(df))$descr
str(npy_decode(npy_encode(df)))
```

A named list is a `.npz`, stored or compressed. `npy_names()` reads the
directory alone, and `names` reads only some members:

```{r}
z <- tempfile(fileext = ".npz")
npy_write(list(weights = x, meta = df), z, compress = TRUE)
npy_names(z)
names(npy_read(z, names = "meta"))
```

## Files from elsewhere

A NumPy file is untrusted input like any other: the header is a Python
literal that NumPy itself evaluates with care. zunpy parses it with a
grammar that knows strings, numbers, tuples, lists and dicts and nothing
else, checks it whole, and compares the size it declares with the bytes
present before allocating anything. A header claiming an array of 2^80
elements costs nothing:

```{r}
hostile <- charToRaw("{'descr': '<f8', 'fortran_order': False, 'shape': (1099511627776, 1099511627776), }")
pad <- (64 - (10 + length(hostile) + 1) %% 64) %% 64
bytes <- c(as.raw(c(0x93, 0x4e, 0x55, 0x4d, 0x50, 0x59, 1, 0)),
           as.raw(c((length(hostile) + pad + 1) %% 256, (length(hostile) + pad + 1) %/% 256)),
           hostile, rep(charToRaw(" "), pad), charToRaw("\n"))
e <- tryCatch(npy_decode(bytes), zunpy_error = function(e) e)
class(e)
```

Every error is a classed condition (see `?"zunpy-conditions"`), so code
can catch a kind of failure rather than match a message. The limits —
`max_size`, `max_header`, `max_dims`, `max_members` — are arguments.

## What does not round-trip

Some things R has and NumPy does not, and the reverse:

- `dimnames` and names are dropped; a factor is written as its labels.
- R has no 0-d array: a NumPy scalar array reads as a length-1 vector and
  writes back as shape `(1,)`.
- `raw` writes as `|u1`, which reads back as integer.
- A `POSIXct` keeps its instant; its time zone becomes UTC.
- `M8[ns]` beyond about 104 days from 1970 is more precise than a double;
  `datetime = "integer64"` keeps the count exactly.
- A record field's title, its padding and a subarray's shape are not kept,
  and a field's dtype follows its column's R type unless `dtype` names it.
